\chapter{Sparse Data}
\section{Definition}

Data are said sparse when it is contains only a few non-null values. That kind of dataset are really common in Machine Learning application and can be an high influence on the computation. 

The sparsity of a dataset is defined by :

\begin{equation}\label{eqn:sparsity}
sparsity = \frac{\text{\# non-null values}}{\text{\# values}}
\end{equation}

Conversely when a dataset has only a few null values, the data are said dense. The density of the dataset is defined by the inverse of the sparsity:
\begin{equation}\label{eqn:density}
density = \frac{1}{\text{sparsity}}
\end{equation}

Using dense methods and data structure with sparse data could have a severe bad impact on the performance
\section{The Frequency of Sparse Data}
Sparsity is a very useful property in Machine Learning. Some algorithms can have fast optimization, fast evaluation of the model, statistical robustess or other computational advantages.
A lot of machine learning application are using sparse dataset such as recommender system, natural language processing algorithm, 


\section{Real application}
In 2008 Netflix launched a contest, the Netflix Grand Prize \cite{netflixgrandprize}, to improve their recommender system model and to increase the accuracy of predictions and published an sample dataset made with the ratings of anonymous Netflix customers. The dataset had more than 100 millions sampled ratings and it contained about $m=480'186$ users and $m=17'770$ movies \cite{Koren091the}. If stored as a dense matrix, it would need to store $8'532'905'220$ values in memory. That corresponds to a sparsity $\cong \frac{100'000'000}{8'532'905'220} = 0.011719338$.

Storing more than 8 trillions 64-bit floating-point numbers needs more than 64 gigabyte of memory which quickly become unmanageable even for the world's fastest supercomputers. That is why it is absolutely necessary to encode the matrix into a sparse format to drastically decrease the amount of memory needed.


\section{What Sparse linear Algebra resolve and optimize}





\section{Formats}
\subsection{Matrices}
\subsubsection{Coordinates Format}
\subsubsection{Compressed Row Format}
\subsubsection{Compressed Column Format}
\subsection{Tensors - Multi-dimensional arrays}
\subsubsection{Coordinates Format}
\subsubsection{Compressed Fiber Format}
